% \documentclass{vldb}

% \usepackage[final,kerning=true,spacing=true,factor=1100,stretch=15,shrink=15]{microtype}
% \usepackage{balance}

% \input{commands}


\documentclass[sigconf, nonacm]{acmart}

%% The following content must be adapted for the final version
% paper-specific
\usepackage{balance}

\newcommand\vldbdoi{XX.XX/XXX.XX}
\newcommand\vldbpages{XXX-XXX}
% issue-specific
\newcommand\vldbvolume{15}
\newcommand\vldbissue{X}
\newcommand\vldbyear{2022}
% should be fine as it is
\newcommand\vldbauthors{\authors}
\newcommand\vldbtitle{\shorttitle} 
% leave empty if no availability url should be set
\newcommand\vldbavailabilityurl{URL_TO_YOUR_ARTIFACTS}
% whether page numbers should be shown or not, use 'plain' for review versions, 'empty' for camera ready
\newcommand\vldbpagestyle{plain}


\input{commands}





\newcommand{\fixme}[1]{\textit{\textcolor{red}{[#1]}}\xspace}











\begin{document}
    % Insert the algorithm

\title{{COSCFair}: Ensuring Subgroup Fairness Through Fair Classification Framework}
% delete the page numbering later
%\fancyfoot[R]{\thepage}

\author{Begum Hattatoglu}
\affiliation{%
  \institution{Utrecht University}  
  \city{Utrecht} 
  \country{Netherlands}   
}
\email{b.hattatoglu@students.uu.nl}

 
\author{Abdulhakim Qahtan}
\affiliation{%
  \institution{Utrecht University}
  \city{Utrecht} 
  \country{Netherlands}
  }
\email{a.a.a.qahtan@uu.nl}



\author{Heysem Kaya}
  \affiliation{%
  \institution{Utrecht University}
  \city{Utrecht} 
  \country{Netherlands} 
 }
 \email{h.kaya@uu.nl}



\author{Yannis Velegrakis}
\affiliation{%
  \institution{Utrecht University and University of Trento}
  \city{Utrecht} 
  \country{Netherlands}}
\email{i.velegrakis@uu.nl}




\begin{abstract}

Machine Learning (ML) algorithms are used in a wide range of applications, which affected societies either directly or indirectly in daily life. ML algorithms are preferred for many tasks that require complex computations with big volumes of data due to the better performance compared to humans. Moreover, people have subjective opinions and points of view, which can lead to bias in their decisions. Unfortunately, ML algorithms are not always objective either. Using ML algorithms in several decision-making systems and other services may cause serious discrimination among some groups of people in society. One of the most significant reasons behind the biased predictions of the algorithms for different demographic groups is the imbalanced representation of each demographic subgroup in the population. We propose COSCFair (Clustering and OverSamling for Fair Classification), which is a framework to ensure fairness among the subgroups without changing the original class labels. COSCFair consists of clustering, oversampling, and classification components, where the classification component considers the outcomes of the clustering algorithm. Our experimental results over different datasets that are widely used as benchmarks to evaluate algorithmic fairness show that our framework yields consistent improvements compared to a set of baseline methods. 


\end{abstract}







\maketitle





%%% do not modify the following VLDB block %%
%%% VLDB block start %%%
\pagestyle{\vldbpagestyle}
\begingroup\small\noindent\raggedright\textbf{PVLDB Reference Format:}\\
\vldbauthors. \vldbtitle. PVLDB, \vldbvolume(\vldbissue): \vldbpages, \vldbyear.\\
\href{https://doi.org/\vldbdoi}{doi:\vldbdoi}
\endgroup
\begingroup
\renewcommand\thefootnote{}\footnote{\noindent
This work is licensed under the Creative Commons BY-NC-ND 4.0 International License. Visit \url{https://creativecommons.org/licenses/by-nc-nd/4.0/} to view a copy of this license. For any use beyond those covered by this license, obtain permission by emailing \href{mailto:info@vldb.org}{info@vldb.org}. Copyright is held by the owner/author(s). Publication rights licensed to the VLDB Endowment. \\
\raggedright Proceedings of the VLDB Endowment, Vol. \vldbvolume, No. \vldbissue\ %
ISSN 2150-8097. \\
\href{https://doi.org/\vldbdoi}{doi:\vldbdoi} \\
}\addtocounter{footnote}{-1}\endgroup
%%% VLDB block end %%%

%%% do not modify the following VLDB block %%
%%% VLDB block start %%%
\ifdefempty{\vldbavailabilityurl}{}{
\vspace{.3cm}
\begingroup\small\noindent\raggedright\textbf{PVLDB Artifact Availability:}\\
The source code, data, and/or other artifacts have been made available at \url{\vldbavailabilityurl}.
\endgroup
}







% \keywords{Algorithmic Fairness, Subgroup Fairness, Clustering, Classification, Oversampling}




\input{Sections/Introduction}
\input{Sections/RelatedWork}
\input{Sections/Problem}
\input{Sections/Theory}

\input{Sections/Framework}

\input{Sections/Evaluation}

\input{Sections/Conclusion}


\bibliographystyle{ACM-Reference-Format}
\bibliography{ref}


\end{document}



